Supplemental material: Ecosystem carbon balance in the Hawaiian Islands under different scenarios of future climate and land use change

### 2021-08-09 Manuscript/journal_correspondence/ERL_responseletter.Rmd

\raggedright
Paul C. Selmants, Ph.D. 
\linebreak
Research Ecologist 
\linebreak
U.S. Geological Survey

## Referee 1, comments to the author:

**The manuscript describes an impressive effort projecting ecosystem carbon balance in the Hawaiian Islands under climate change and land use change scenarios. This assessment of future regional terrestrial carbon dynamics can provide a support to local planning of climate mitigation strategies through ecosystem carbon sequestration. The manuscript is well written and organized. However, more clarification and justification are needed for methodology and assumptions about how future NPP and land cover change is modelled. And also, more details are needed such as maps illustrating how projected change in carbon flux (e.g., NPP and NBP) and carbon pools (live biomass, soil organic soil) is spatially distributed. I have provided detailed comments below.**

**L109, why need to assign each cell a state type? What is consideration for it?**  The state of each cell tracks a range of information about that cell, including land cover class, moisture zone, island, age, and time since transition. This set of information is used to define the probabilities that each cell can undergo a particular type of transition, as well as how often each of these transitions can occur. We have added text and references to this section and the next that better describes the state of each cell (lines xxx).

**L114, transitions between ‘state type’? Is it ‘land cover classes’? According to the definition of state type as unique combination of moisture zones, islands and land cover classes at L110, is there transition from dry forest and mesic forest in the same island?**  
State type in this model is the unique combination of moisture zone, island, and land cover class, as stated on line 110 of the original manuscript. This information affects transition probabilities, such that dry grassland on Hawaiʻi Island has a different probability of transitioning to agriculture than mesic shrubland on Oʻahu. We have added text that better defines how state type influences transition probabilities ().

**L123-126, high land use scenario samples historical rate between median and maximum, but why not low land use scenario sample rates between minimum and median? What is the justification that low land use scenario will have rate lower than historical records?**  
To distinguish the impact of a low land use scenario from a high land use scenario on changes in agricultural land area, we sampled rates of agricultural contraction and agricultural expansion from a uniform distribution bounded by zero and the minimum historical rate. Setting the lower boundary of agricultural contraction and expansion rates at zero in the low land use scenario allows for the possibility that agricultural land area will stabilize over time. We consider this to be a reasonable possibility under a low land use scenario, especially given the volatility in agricultural land area change experienced in the Hawaiian Islands over the recent historical period (1992-2010). We have added text that justifies bounding  agricultural land area change rates between zero rather and the minimum 1992-2010 historical rates in the low land use scenario (lines xxx - xxx).

@@

**L184, what is the temporal coverage and resolution of these climate projections? Do them only cover the period of mid-century (2049-2069) and end-of-century (2070-2099) as described in L179?**  
Yes, climate projections only cover the period of mid-century (2049-2069) and end-of-century (2070-2099) as described on line 179 of the original manuscript. We have added text that describes how to the supplementary climate section that provides more detail on climate projections (lines xxx) No changes were made in response to this comment.

**L239: is the NPP decrease during 2010-2020 evident in satellite observations? To what extent of this decrease is caused by climate conditions change rather than by forest area changes?**  
Statewide NPP declined by approximately 2% on average between 2010 and 2020 across all scenarios and Monte Carlo iterations (8.13 Tg y^-1^ to 7.95 Tg y^-1^), and the ranges of statewide NPP estimates for each of these years broadly overlapped (2010 range = 7.96 - 8.28 Tg ^y-1^, 2020 range = 7.85 - 8.1 Tg ^y-1^). This modeled slight decline in NPP over 10 years, which is well within the error range of coarse resolution satellite observations, was due almost entirely to land cover change - which includes but is not limited to forest area change. We have added text that describes this change in NPP over the contemporary (2010-2020) time period (lines xxx-xxx).

**L242. It is misleading to state ‘climate change led to increased Rh over time’ as figure 3 shows decreasing trend of Rh under high land use scenario and two RCP climate change scenarios. It should be revised as something like ‘switching from RCP 4.5 to 8.5 led to increased Rh’.**
We agree, this phrasing was misleading. We have altered the text

**L244, why the increase in total soil carbon stock (figure 2) will not enhance Rh? Is it due to the assumption made by this work that Rh is only ratio of NPP and latter decrease over time? What are drivers of Rh decrease?**  
Soil carbon stocks increase consistently only under the RCP 4.5 radiative forcing scenario. Under RCP 8.5, soil carbon stocks reach an asymptote by 2080 in the low land use scenario and begin to decline slightly in the high land use scenario. Also, Rh only declines consistently under the high land use scenario (Figure 3). In the low land use scenario Rh remains relatively stable after 2060 under RCP 4.5 and increases after 2020 under RCP 8.5. As stated previously, NPP is a direct driver of Rh because the vast majority of carbon respired heterotrphically from terrestrial ecosystems was fixed by photosynthesis that same year (Kuzyakov et al. 2010; see lines xxx of revised manuscript). The drivers of Rh are a combination of climate change and land use change. High rates of urbanization remove plants and compress and pave over soils, reducing carbon input and lowering microbial activity. Climate change on the the other hand, has a ʻpush-pullʻ impact on Rh. On the one hand

**L244, why Rh shows very sharp decrease before 2020 under all land use scenarios with all RCPs, such decrease rate seems to be higher than the rest period. Are these decreases supported by observations?**  
Land use change rates are the same in all scenarios from 2010-2020, which explains why Rh behaves consistently across these scenarios. We have added text to make that clear in the Methods section (lines xxx). The decrease in Rh before 2020 is driven almost entirely by land use change, specifically the increase in urban area at the expense of agricultural land (supplementary Figure 5). Agriculture has a higher rate of Rh per unit of NPP than other land cover classes, and so Rh declined more than overall NPP as agricultural land was converted to urban area. We have added text that describes these trends (lines xxx).

@@

**I suggest that authors check the bibliographic references in the main text and revise the format of documents to meet the requirement of the Journal.**
We are using the Institute of Physics (IOP) citation style language (csl), which meets the guidelines for this Journal. No changes were made in response to this comment.

**In the abstract, please add more information about the methods of the experiment.**  We have added text to the abstract that provides more information about the methods of this project (lines xxx).

**In the abstract, please don't write the same conclusion lines as they were written in the main conclusion section.**  
We have altered the text so that the conclusions in the abstract are not written identically to those in the Conclusions section (lines xxx).

**Hypothesis should be given. How this work is different from the available literature?**
A hypothesis would be warranted if the purpose of this study was to test existing theory using a controlled experiment, which was not the case. The purpose of this study was to use existing theory to make projections about the future, and so we believe a statement of objectives is sufficient. We describe how this work compares to the available literature in the Discussion section (e.g., lines xxx), as recommended by the author guidelines for this Journal. No changes were made in response to this comment.

@@

**Objectives should be rewritten in detail and comprehensively.** 
We believe we have stated the objectives of this study clearly and succinctly, both in the abstract (lines xxx) and at the end of the Introduction section (lines xxx). That said, we have altered the text to add more detail to our stated objectives (lines xxx-xxx).

**How did the authors perform the calibration and validation process?**
We describe calibration of the model in the Methods section on lines xxx and lines xxx. Validation of ecosystem carbon balance estimates in the Hawaiian Islands is

**How did they select the sensitive parameters of the model that they used?**
We are unsure exactly what the referee is referrring to in this question, as they do not point to a particular portion of the manuscript. There are a number of parameters in this model, and we describe in detail how they were selected in both the Methods section and the Supplementary section. Without further clarification, we do not know how to better respond to this question. No changes were made in response to this comment.

### 2021-08-09 Manuscript/HI_Model_ms.Rmd
updated references, manuscript file, and figure files

## 2. Methods
We used the Land Use and Carbon Scenario Simulator (LUCAS), an integrated landscape change and carbon gain-loss model, to project changes in ecosystem carbon balance for the seven main Hawaiian Islands. The landscape change portion of LUCAS is a state-and-transition model, where 'state' is the pre-defined information specific to each simulation cell (e.g., land cover class, climate zone) and 'transition' is the change in state of each simulation cell over time [@daniel_state-and-transition_2016]. Landscape change occurs by applying a Monte Carlo approach to track the state type and age of each simulation cell in response to a pre-determined set of transition pathways [@daniel_state-and-transition_2016]. Transition pathways define which state types can be converted to any other state type. The carbon gain-loss portion tracks carbon stocks within each simulation cell over time as continuous state variables, along with a pre-defined set of continuous flows specifying rates of change in stock levels over time [@daniel_integrating_2018; @sleeter_effects_2019]. We parameterized the Hawaiʻi LUCAS model to estimate annual changes in carbon stocks and fluxes in response to land use, land use change, wildland fire, and long-term climate variability for the time period 2010-2100.

### *2.1 Study area*
The spatial extent of this study was the terrestrial portion of the seven main Hawaiian Islands (figure 1), a total land area of 16,554 km^2^. We subdivided this landscape into a grid of 264,870 simulation cells, each of which was 250 x 250 m in size. Each simulation cell was assigned to one of 210 possible state types based on the unique combination of three moisture zones (dry, mesic, and wet; supplemental figure 1), seven islands, and ten discrete land cover classes (figure 1).

@@

### *2.2 States and transitions*
We developed two land use scenarios (low and high) with transition pathways modified from @daniel_state-and-transition_2016. Transition pathways were pre-defined to represent urbanization, agricultural contraction, agricultural expansion, harvesting of tree plantations, and wildfire. Agriculture, forest, grassland, shrubland, and tree plantation land cover classes each had multiple transition pathways, i.e., they could each be converted into multiple other land cover classes. The barren land cover class could only transition to developed, so its sole transition pathway was urbanization. Although most land cover classes had an urbanization transition pathway, there was no transition pathway out of an urbanized (developed) state. Water and wetland land cover classes remained static throughout the simulation period.

 Transition targets set the area to be transitioned over time. These targets were based on historical trends of land use change in the Hawaiian Islands from 1992-2011 [@noaa_coastal_2020] and on population projections for the State of Hawaiʻi [@kim_population_2018]. For the high land use scenario, transition rates for each timestep and Monte Carlo realization were sampled from uniform distributions bounded by the median and maximum historical rates of agricultural contraction, agricultural expansion, and urbanization for each island. For the low land use scenario, rates of agricultural contraction and expansion were sampled from uniform distributions bounded by zero and the minimum historical rates for each island. Urbanization rates in the low land use scenario were based on island-level population estimates and projections at five year intervals from 2010-2045 [@kim_population_2018]. We converted population projections into urbanization transition targets following @sleeter_future_2017 by calculating population density for each island and then projecting future developed area based on the five-year incremental change in island population. The spatial extent of agricultural contraction, agricultural expansion, and urbanization was constrained in both land use scenarios based on existing zoning maps [@daniel_state-and-transition_2016]. Tree plantation forestry in the State of Hawaiʻi is primarily short-rotation (5-7 years) *Eucalyptus* spp., harvested at a rate of approximately 2 km^2^ y^-1^ over the past decade. Transition targets for tree plantation harvest were set at ~75% of recent historical rates in the high land use scenario and ~40% of recent historical rates in the low land use scenario [@daniel_state-and-transition_2016]. In both land use scenarios, approximately 60% of tree plantation harvests were replacement harvests resulting in conversion to agriculture. The remaining 40% were rotation harvests replanted to *Eucalyptus* spp. No land cover class could be converted to tree plantation in either land use scenario, consistent with recent historical trends of stable or declining area of tree plantations.

The wildfire transition sub-model was modified from @daniel_state-and-transition_2016 by incorporating a new 21-year historical wildfire spatial database of the Hawaiian Islands (supplemental figure 2). We used this new spatial database to calculate historical wildfire size distribution and ignition probabilities for each unique combination of moisture zone (supplemental figure 1), island, and land cover class (figure 1) for the years 1999-2019. Starting in 2020, the number and size of fires was randomly drawn from one of these historical year-sets for each timestep and Monte Carlo realization, using burn severity probabilities from @selmants_baseline_2017. Wildfire in the low land use scenario was sampled from the subset of historical fire years at or below the median area burned statewide from 1999-2019. The high land use scenario sampled from historical fire years above the median area burned over the same 21-year period (supplemental figure 2a). The vast majority of wildland fire in Hawaiʻi is human-caused and initiates in non-native grasslands and shrublands [@trauernicht_contemporary_2015], but can spread into nearby forest areas. Moderate to high severity fires that spread into mesic and wet forests can convert these areas into grasslands about half the time. However, moderate to high severity fires in mesic and wet forests account for <5% of the total annual area burned, on average [@selmants_baseline_2017].

@@

 Initial carbon stocks and baseline carbon flows were estimated based on the state type and age of each simulation cell using a lookup table derived from the Integrated Biosphere Simulator [IBIS; @foley_integrated_1996; @liu_critical_2020], a process-based dynamic global vegetation model. We initiated IBIS with minimal vegetation and simulated forward for 110 years using 30-year climate normals for the Hawaiian Islands [@giambelluca_online_2013; @giambelluca_evapotranspiration_2014]. We calibrated IBIS carbon stocks with statewide gridded datasets of soil organic carbon [@ssurgo_2016] and forest aboveground live biomass [@asner_rapid_2016]. We also calibrated gross photosynthesis in IBIS using a Hawaiʻi-specific gridded dataset derived from MODIS satellite imagery [@kimball_evaluating_2017].

 Net primary production for each simulation cell was calculated as the mean IBIS-derived value for each combination of moisture zone and land cover class (supplemental Table 1) adjusted with a spatially explicit stationary growth multiplier to reflect local variation driven by microclimate [supplemental Figure 4; @sleeter_effects_2019]. We calculated this spatial growth multiplier as the NPP anomaly for each simulation cell relative to mean NPP values for each combination of moisture zone and land cover class based on empirical relationships between total annual NPP and mean annual rainfall or temperature [@schuur_productivity_2003; @del_grosso_global_2008] using Hawaiʻi-specific climate data [@giambelluca_online_2013; @giambelluca_evapotranspiration_2014]. Soil carbon flux to the atmosphere (R~h~) and aquatic soil carbon losses (leaching and overland flow) were estimated as the ratio of the IBIS-derived flux for each combination of moisture zone and state type to the microclimate-adjusted NPP value for each simulation cell. All other carbon flow rates were estimated as the ratio of the mean IBIS-derived flux for each combination of moisture zone and state type to the size of the originating carbon stock at each age [@sleeter_effects_2018; @daniel_integrating_2018].

>>>>

## 2. Methods
We used the Land Use and Carbon Scenario Simulator (LUCAS), an integrated landscape change and carbon gain-loss model, to project changes in ecosystem carbon balance for the seven main Hawaiian Islands. The landscape change portion of LUCAS is a state-and-transition model, where 'state' is the pre-defined information specific to each simulation cell (e.g., land cover class, climate zone) and 'transition' is the change in state of each simulation cell over time [@daniel_state-and-transition_2016]. Landscape change occurs by applying a Monte Carlo approach to track the state and age of each simulation cell in response to a pre-determined set of transition pathways [@daniel_state-and-transition_2016]. Transition pathways define which state types can be converted to any other state type. The carbon gain-loss portion tracks carbon stocks within each simulation cell over time as continuous state variables, along with a pre-defined set of continuous flows specifying rates of change in stock levels over time [@daniel_integrating_2018; @sleeter_effects_2019]. We parameterized the Hawaiʻi LUCAS model to estimate annual changes in carbon stocks and fluxes in response to land use, land use change, wildland fire, and long-term climate variability for the time period 2010-2100.

### *2.1 Study area*
The spatial extent of this study was the terrestrial portion of the seven main Hawaiian Islands (figure 1), a total land area of 16,554 km^2^. We subdivided this landscape into a grid of 264,870 simulation cells, each of which was 250 x 250 m in size. Each simulation cell was assigned to one of 210 possible state types based on the unique combination of three moisture zones (dry, mesic, and wet; supplemental figure 1), seven islands, and ten discrete land cover classes (figure 1). We also tracked the age and time since transition for each simulation cell as continuous state variables [@daniel_state-and-transition_2016].

@@

### *2.2 States and transitions*
We developed two land use scenarios (low and high) with transition pathways modified from @daniel_state-and-transition_2016. Transition pathways were pre-defined to represent urbanization, agricultural contraction, agricultural expansion, harvesting of tree plantations, and wildfire. Agriculture, forest, grassland, shrubland, and tree plantation land cover classes each had multiple transition pathways, i.e., they could each be converted into a variety of other land cover classes. The barren land cover class could only transition to developed, so its sole transition pathway was urbanization. Although most land cover classes had an urbanization transition pathway, there was no transition pathway out of an urbanized (developed) state. Water and wetland land cover classes remained static throughout the simulation period.

Transition targets set the area to be transitioned over time. These targets were based on recent historical trends of land use change in the Hawaiian Islands from 1992-2011 [@noaa_coastal_2020] and on population projections for the State of Hawaiʻi [@kim_population_2018]. For the high land use scenario, transition rates for each timestep and Monte Carlo realization were sampled from uniform distributions bounded by the median and maximum recent historical rates of agricultural contraction, agricultural expansion, and urbanization for each island. For the low land use scenario, rates of agricultural contraction and expansion were sampled from uniform distributions bounded by zero and the minimum recent historical rates for each island. Bounding agricultural change rates between zero and the recent historical minimum shifts the balance toward agricultural contraction over time in the low land use scenario, which reduces the land area under active cultivation and contrasts with the high variability in agricultural land change rates during the recent historical past [@suryanata_diversified_2009]. Urbanization rates in the low land use scenario were based on island-level population estimates and projections at five year intervals from 2010-2045 [@kim_population_2018]. We converted population projections into urbanization transition targets following @sleeter_future_2017 by calculating population density for each island and then projecting future developed area based on the five-year incremental change in island population. The spatial extent of agricultural contraction, agricultural expansion, and urbanization was constrained in both land use scenarios based on existing zoning maps [@daniel_state-and-transition_2016]. Transition targets for tree plantation harvest were set at ~75% of recent historical rates in the high land use scenario and ~40% of recent historical rates in the low land use scenario [@daniel_state-and-transition_2016]. Tree plantation forestry in the State of Hawaiʻi is primarily short-rotation (5-7 years) *Eucalyptus* spp., harvested at a rate of approximately 2 km^2^ y^-1^ over the past decade. In both the high and low land use scenarios, approximately 60% of tree plantation harvests were replacement harvests resulting in conversion to grassland or agriculture. The remaining 40% were rotation harvests replanted to *Eucalyptus* spp. There was no transition pathway from any land cover class into tree plantation in either land use scenario, which is consistent with recent historical trends of stable or declining tree plantation land area.

The wildfire transition sub-model was modified from @daniel_state-and-transition_2016 by incorporating a new 21-year historical wildfire spatial database of the Hawaiian Islands (supplemental figure 2). We used this new spatial database to calculate historical wildfire size distribution and ignition probabilities for each unique combination of moisture zone (supplemental figure 1), island, and land cover class (figure 1) for the years 1999-2019. Starting in 2020, the number and size of fires was randomly drawn from one of these historical year-sets for each timestep and Monte Carlo realization, using burn severity probabilities from @selmants_baseline_2017. Wildfire in the low land use scenario was sampled from the subset of historical fire years at or below the median area burned statewide from 1999-2019. The high land use scenario sampled from historical fire years above the median area burned over the same 21-year period (supplemental figure 2a). The vast majority of wildland fire in Hawaiʻi is human-caused and initiates in non-native grasslands and shrublands [@trauernicht_contemporary_2015], but can spread into nearby forest areas. Moderate to high severity fires that spread into mesic and wet forests can convert these areas into grasslands about half the time. However, moderate to high severity fires in mesic and wet forests account for < 5% of the total annual area burned, on average [@selmants_baseline_2017; @daniel_integrating_2018].

@@

Initial carbon stocks and baseline carbon flows were derived from the Integrated Biosphere Simulator [IBIS; @foley_integrated_1996; @liu_critical_2020], a process-based dynamic global vegetation model. We initiated IBIS with minimal vegetation and simulated forward for 110 years using 30-year climate normals for the Hawaiian Islands [supplemental Figure 3; @giambelluca_online_2013; @giambelluca_evapotranspiration_2014]. We calibrated IBIS carbon stocks with statewide gridded datasets of soil organic carbon [@ssurgo_2016] and forest aboveground live biomass [@asner_rapid_2016] converted to total forest live biomass by applying a power function to calculate the belowground portion [@mokany_critical_2006] as described in @selmants_baseline_2017. We calibrated gross photosynthesis in IBIS using a Hawaiʻi-specific gridded dataset of gross primary production (GPP) derived from MODIS satellite imagery [@kimball_evaluating_2017]. Mean annual GPP from @kimball_evaluating_2017 ranged from 0.123 - 3.88 kg C m^-2^ y^-1^ across the Hawaiian Islands, resulting in a NPP-to-GPP ratio ranging from 0.37 - 0.56.

We initiated net primary production for each simulation cell in LUCAS as the mean IBIS-derived value for each combination of moisture zone and land cover class (supplemental Table 1) adjusted with a set of spatially explicit stationary NPP multipliers to reflect local variation driven by microclimate [supplemental Figure 4; @sleeter_effects_2018; @sleeter_effects_2019]. The spatial NPP multipliers were the NPP anomaly for each simulation cell relative to mean empirically-derived NPP values for each state type (supplemental Figure 4). Soil carbon flux to the atmosphere (R~h~) and aquatic soil carbon losses (leaching and overland flow) were estimated as the ratio of the IBIS-derived flux for each state type to the microclimate-adjusted NPP value for each simulation cell. All other carbon flow rates were estimated as the ratio of the mean IBIS-derived flux for each state type to the size of the originating carbon stock at each age [@sleeter_effects_2018; @daniel_integrating_2018].

### 2021-08-09 Manuscript/HI_Model_supp.Rmd
updated references, manuscript file, and figure files

State Classes for Agriculture, Forest, Grassland, Shrubland, and Tree Plantations were stratified into three Moisture Zones - Dry, Mesic, and Wet (Figure S1). These three zones were based on a moisture availability index (MAI), calculated as mean annual precipitation (MAP) minus potential evapotranspiration [PET; @price_mz_2012]. Areas where where MAI values were less than zero (i.e., where MAP < PET) were classified in the Dry Moisture Zone. Areas with MAI values between zero and 1,661 were classified in the Mesic Moisture Zone. The MAI value of 1,661 is roughly equivalent to areas at 1,000 m elevation that receive 2,500 mm of annual rainfall [@price_mz_2012]. Areas with MAI values greater than 1,661 were classified in the Wet Moisture Zone.

```{r fig.align = 'center', out.width = "95%", fig.cap = "Supplemental - Moisture zones of the seven main Hawaiian Islands, adapted from Jacobi \\textit{et al.} (2017)."}
knitr::include_graphics("./fig_images/figS1_MZ.png")
```

@@

This study used a new spatial database of wildland fire perimeters on the main Hawaiian Islands from 1999-2019 to calculate annual area burned (Figure S2), wildland fire probabilities by Moisture Zone and State Class, and wildland fire size distributions. This new database compiles prior mapping efforts and data collections with new fire perimeter data mapped directly by Dr. Clay Trauernicht (Department of Natural Resources and Environmental Management, University of Hawaiʻi at Mānoa). The goal of this effort was to locate and map all fires greater than or equal to 20 hectares, but some smaller fires were included as detected in imagery. Fire perimeters for the years 2002-2011 were primarily from the U.S. Geological Survey's Monitoring Trends in Burn Severity (MTBS; https://www.mtbs.gov). The Hawaii Wildfire Management Organization (www.hawaiiwildfire.org) provided ground-based, GPS-mapped fire perimeters from Hawaiʻi Island, primarily in the Kona and Kohala regions. The U.S. National Park Service provided ground-based, GPS-mapped fire records from Hawaiʻi Volcanoes National Park. The Oʻahu Army Natural Resource Program provided ground-based, GPS-mapped fire records on Oʻahu. All other fires were mapped directly by Dr. Trauernicht using LANDSAT and Sentinel-2 satellite imagery. Data from USGS MTBS were prioritized for the years 2002-2011 in the case of duplicate records.

```{r fig.align = 'center', out.width = "99%", fig.cap = "Supplemental - Annual area burned by wildland fire in the State of Hawaiʻi from 1999-2019, summed by year across the seven main Hawaiian Islands (A) and within each of the four largest islands (B). The dashed horizontal line in (A) represents the median area burned from 1999-2019."}
knitr::include_graphics("./fig_images/figS2_AnnualAreaBurned.png")
```

@@

Spatially explicit contemporary mean annual temperature and rainfall data for the main Hawaiian Islands at 250-m resolution (Figure S3) are from @giambelluca_online_2013 and @giambelluca_evapotranspiration_2014. Spatially explicit projections of change in annual temperature and change in annual rainfall by the end of century (2100; Figure S4) were developed using statistically downscaled CMIP5 climate projections for the Hawaiian Islands from @timm_statistical_2015 and @timm_future_2017 under RCP 4.5 and RCP 8.5.

```{r fig.align = 'center', out.width = "95%", fig.cap = "Supplemental - Mean annual temperature (top panel) and mean annual rainfall (bottom panel) for the seven main Hawaiian Islands. Data from Giambelluca \\textit{et al.} (2013) and Giambelluca \\textit{et al.} (2014)."}
knitr::include_graphics("./fig_images/figS3_CurrentClimate.png")
```
```{r fig.align = 'center', out.width = "99%", fig.cap = "Supplemental - Projected change in mean annual temperature (top panels) and mean annual rainfall (bottom panels) by 2100 under RCP 4.5 and RCP 8.5 based on statistical downscaling of an ensemble of climate models (CMIP5)."}
knitr::include_graphics("./fig_images/figS4_FutureClimate.png")
```

@@

Projections of total land area covered by each of five State Classes (Agriculture, Developed, Forest, Grassland and Shrubland) were summed by year and Monte Carlo iteration across the seven main Hawaiian Islands for each of two land use scenarios (low and high; Figure S5).

```{r fig.align = 'center', out.width = "99%", fig.cap = "Supplemental - Projections of total land area by year for each of five State Classes in the seven main Hawaiian Islands (Agriculture, Developed, Forest, Grassland, and Shrubland) under low and high land use change scenarios for the period 2010-2100. Solid lines represent the mean of 30 Monte Carlo realizations and shaded areas represent minimum and maximum Monte Carlo values."}
knitr::include_graphics("./fig_images/figS5_StateClassArea.png")
```

>>>>

Agriculture, Forest, Grassland, Shrubland, and Tree Plantation land cover classes were stratified into three Moisture Zones - Dry, Mesic, and Wet (supplemental Figure 1). These three zones were based on a moisture availability index (MAI), calculated as mean annual precipitation (MAP) minus potential evapotranspiration [PET; @price_mz_2012]. Areas where where MAI values were less than zero (i.e., where MAP < PET) were classified in the Dry Moisture Zone. Areas with MAI values between zero and 1,661 were classified in the Mesic Moisture Zone. The MAI value of 1,661 is roughly equivalent to areas at 1,000 m elevation that receive 2,500 mm of annual rainfall [@price_mz_2012]. Areas with MAI values greater than 1,661 were classified in the Wet Moisture Zone.
```{r fig.align = 'center', out.width = "99%", fig.cap = "Supplemental - Moisture zones of the seven main Hawaiian Islands, adapted from Jacobi \\textit{et al.} (2017)."}
knitr::include_graphics("./fig_images/figS1_MZ.png")
```

@@

We used a new spatial database of wildland fire perimeters on the main Hawaiian Islands from 1999-2019 to calculate annual area burned (supplemental Figure 2), wildland fire probabilities by state type, and wildland fire size distributions. This new database compiles prior mapping efforts and data collections with newly mapped fire perimeter data from Dr. Clay Trauernicht (Department of Natural Resources and Environmental Management, University of Hawaiʻi at Mānoa). The goal was to locate and map all fires greater than or equal to 20 hectares, but some smaller fires were included as detected in imagery. Fire perimeters for the years 2002-2011 were primarily from the U.S. Geological Survey's Monitoring Trends in Burn Severity (MTBS; https://www.mtbs.gov). The Hawaii Wildfire Management Organization (www.hawaiiwildfire.org) provided ground-based, GPS-mapped fire perimeters from Hawaiʻi Island, mostly from the Kona and Kohala regions. The U.S. National Park Service provided ground-based, GPS-mapped fire records from Hawaiʻi Volcanoes National Park. The Oʻahu Army Natural Resource Program provided ground-based, GPS-mapped fire records on Oʻahu. All other fires were mapped directly by Dr. Trauernicht using LANDSAT and Sentinel-2 satellite imagery. Data from USGS MTBS were prioritized for the years 2002-2011 in the case of duplicate records.
```{r fig.align = 'center', out.width = "95%", fig.cap = "Supplemental - Annual area burned by wildland fire in the State of Hawaiʻi from 1999-2019, summed by year across the seven main Hawaiian Islands (A) and within each of the four largest islands (B). The dashed horizontal line in (A) represents the median area burned from 1999-2019."}
knitr::include_graphics("./fig_images/figS2_AnnualAreaBurned.png")
```

@@

Spatially explicit contemporary mean annual temperature and rainfall data for the main Hawaiian Islands at 250-m resolution (supplemental Figure 3) are from @giambelluca_online_2013 and @giambelluca_evapotranspiration_2014. Spatially explicit projections of change in annual temperature and annual rainfall for mid-century (2049-2069) and end of century (2070-2099; supplemental Figure 4) are  from statistically downscaled CMIP5 climate projections under RCP 4.5 and RCP 8.5 from @timm_statistical_2015 and @timm_future_2017. To avoid biases, the modeled present-day (1975–2005) CMIP5 climatology was used to standardize the resulting predictor time series, and the simulated future changes were measured relative to their present-day mean states. See @timm_statistical_2015 and @timm_future_2017 for a more detailed description of potential statistical downscaling biases.  
```{r fig.align = 'center', out.width = "80%", fig.cap = "Supplemental - Contemporary 30-year climate normals for mean annual temperature (top panel) and mean annual rainfall (bottom panel) for the seven main Hawaiian Islands. Data from Giambelluca \\textit{et al.} (2013) and Giambelluca \\textit{et al.} (2014)."}
knitr::include_graphics("./fig_images/figS3_CurrentClimate.png")
```
```{r fig.align = 'center', out.width = "95%", fig.cap = "Supplemental - Projected change in mean annual temperature (top panels) and mean annual rainfall (bottom panels) by 2100 under RCP 4.5 and RCP 8.5 based on statistical downscaling of an ensemble of climate models (CMIP5)."}
knitr::include_graphics("./fig_images/figS4_FutureClimate.png")
```
### Initializing NPP 
Initial values for net primary production (NPP) in LUCAS were calculated using a lookup table derived from the Integrated Biosphere Simulator (IBIS; supplemental Table 1). These values were adjusted with spatially explicit stationary NPP multipliers to reflect local variation driven by microclimate (supplemental Figure 4) as described in detail in @sleeter_effects_2018 and @sleeter_effects_2019. Briefly, we derived the set of NPP spatial multipliers by first calculating NPP independently of the IBIS values for each simulation cell using empirical relationships between total annual NPP and mean annual rainfall or temperature [@schuur_productivity_2003; @del_grosso_global_2008] based on Hawaiʻi-specific climate data (supplemental Figure 3; @giambelluca_online_2013; @giambelluca_evapotranspiration_2014). We then calculated the NPP spatial multipliers as the NPP anomaly for each simulation cell relative to the mean of these empirically-derived NPP values for each state type (i.e., the unique combination of moisture zone and land cover class; supplemental Figure 4). These NPP multipliers were then applied to the IBIS-derived values to create a spatial distribution of NPP used to initiate the LUCAS model (supplemental Figure 4). 
```{r}
library(kableExtra)
ibisnpp <- read.csv("IBISnpp_wide.csv", stringsAsFactors = FALSE)
kbl(ibisnpp, booktabs = T, caption = "Supplemental - Mean IBIS-derived NPP values by Land Cover and Moisture Zone, in kg of C per square meter per year.") %>%
   add_header_above(c(" ", "NPP by Moisture Zone" = 3))
```
```{r fig.align = 'center', out.width = "80%", fig.cap = "Supplemental - Stationary spatial NPP multipliers (top panel) and initial NPP based on the product of mean IBIS-derived NPP values (supplemental Table 1) and stationary spatial NPP multipliers (above). Developed and Barren land cover types are excluded from both maps."}
knitr::include_graphics("./fig_images/figS5_initialNPP.png")
```

@@

Projections of total land area covered by each of five State Classes (Agriculture, Developed, Forest, Grassland and Shrubland) were summed by year and Monte Carlo iteration across the seven main Hawaiian Islands for each of two land use scenarios (low and high; supplemental Figure 5). The overall decline in Agricultural land between 2020 and 2040 in the low land use scenario led directly to the increase in Forest, Grassland, and Shrubland area over the same time period. Although Agricultural land area in the high land use scenario appears more stable over time, the trend actually reflects a balance between high rates of agricultural contraction in some areas and high rates of agricultural expansion in others.     
```{r fig.align = 'center', out.width = "99%", fig.cap = "Supplemental - Projections of total land area by year for each of five State Classes in the seven main Hawaiian Islands (Agriculture, Developed, Forest, Grassland, and Shrubland) under low and high land use change scenarios for the period 2010-2100. Solid lines represent the mean of 30 Monte Carlo realizations and shaded areas represent minimum and maximum Monte Carlo values."}
knitr::include_graphics("./fig_images/figS6_StateClassArea.png")
```
